%==============================================================================
\section{DeltaSort Algorithm}
\label{sec:algorithm}
%==============================================================================

DeltaSort is an insertion-based update-aware sorting algorithm. Below, we first give an overview of the algorithm, then formally prove its correctness, and finally analyze its expected complexity under a random update model.

\subsection{Overview}

DeltaSort operates in two phases (see~\algoref{alg:deltasort} for pseudocode and~\figref{fig:delta-sort-example} for an example). The first phase establishes a structure among updated values, and the second phase exploits this structure to efficiently sort the array.

\paragraph{Phase 1: Establish structure} DeltaSort begins by extracting the values at the updated indices, sorting them (e.g., via indirect heapsort using the index list as indirection), and writing them back into the array in increasing index order. This step enforces order among updated positions: if $i < j$ are both updated indices, then $A[i] \le A[j]$ (according to \texttt{cmp}) after Phase~1. The unchanged values are already in order by design. Hence, any remaining order violations must occur only between updated values and their unchanged neighbors.

\paragraph{Phase 2: Fix violations}
With the above structure in place, DeltaSort fixes the array one updated index at a time left to right. When an updated index is first encountered, we first determine its \emph{direction}:

\input{\figdir/pseudocode}

\begin{definition}[Direction]
\label{def:direction}
During Phase~2, for an index $i \in U$, we define its \emph{direction} based on its local order with its left neighbor:
\begin{enumerate}
  \item \textbf{LEFT ($L$)}: Value is ordered incorrectly to its left neighbor and hence \emph{must} move left: $i > 0 \text{ and } A[i-1] > A[i]$
  \item \textbf{RIGHT ($R$)}: Value is ordered correctly to its left neighbor or it is the first value and hence \emph{may} move right: $i = 0 \text{ or } A[i-1] \le A[i]$
\end{enumerate}
\end{definition}

A few important things to note about this definition:
\begin{itemize}
  \item Direction captures on which side of its current position the target of an updated index lies. Hence, direction is not defined (or needed) for unchanged indices ($i \notin U$).
  \item The definition of direction is asymmetric. $L$ requires a strict violation ($A[i-1] > A[i]$), while $R$ includes both violating and non-violating cases, as it may or may not violate order with its right neighbor. This is deliberate because DeltaSort treats all $R$ values identically, so combining these cases simplifies the implementation (see~\algoref{alg:deltasort}).
\end{itemize}

\input{\figdir/delta-sort-example}

Phase~2 (lines~9--29) processes updated indices from left to right while maintaining a growing sorted prefix. We describe Phase~2 by first establishing a loop invariant and then proving its correctness by showing that the loop invariant holds at initialization, is preserved during maintenance, and that the array is fully sorted at termination.

\paragraph{State} We utilize the following state information during Phase~2:
\begin{itemize}
  \item \texttt{updatedIndices}: Sorted list of updated indices with a sentinel $n$ appended at the end.
  \item $\ell$: The boundary of the sorted prefix, $0$ at the start.
  \item \texttt{pendingRight}: Stack of pending $R$ indices, empty at the start.
\end{itemize}

\paragraph{Loop invariant} At the start of each iteration of the loop at line~10, the following conditions hold for the current index $i$:
\begin{enumerate}
  \item The subarray $A[0..\ell-1]$ is sorted.
  \item There are no $L$ violations in the range $[\ell, i-1]$.
  \item The stack contains pending $R$ indices within the range $[\ell, i-1]$.
  \item If the previous updated index had $L$ direction, then \texttt{pendingRight} is empty.
\end{enumerate}

\paragraph{Initialization} The invariant trivially holds before the first iteration ($i = \texttt{updatedIndices}[0]$).

\paragraph{Maintenance} We will now show that if the invariant holds for $i$, it also holds for the next updated index. Let $\texttt{dir}$ denote the direction of $i$.

\textbf{Case 1: $\texttt{dir} = L$}
\begin{enumerate}
  \item We first set $r \gets i - 1$. This marks the right boundary beyond which the $R$ values in \texttt{pendingRight} cannot move, because after Phase~1, updated values are ordered and hence cannot cross each other in either direction.
  \item We then flush all pending $R$ indices $j$ in LIFO order. Each $R$ index $j$ is fixed by binary insertion within the range $[j+1, r]$, where $r$ is advanced leftward after each fix to ensure that we search within a minimal range for maximum efficiency. By construction, the range $[j+1, r]$ contains no violations, so binary search finds the correct insertion point. This step preserves conditions (3) and (4) of the loop invariant.
  \item After all $R$s are fixed, we fix the current $L$ index $i$ by binary insertion within $[\ell, i-1]$. This range also contains no violations because all $R$ violations are already fixed in the step above, and by invariant condition (2), this range contained no $L$ violations to begin with. Finally, we set $\ell \gets$ (insertion point of $i$) $+ 1$ (not needed for correctness, but reduces the search range). This preserves conditions (1) and (2) of the loop invariant.
\end{enumerate}

\textbf{Case 2: $\texttt{dir} = R$}
\begin{enumerate}
  \item If \texttt{pendingRight} is empty, this is the first $R$ in a new segment, so we set $\ell \gets i$. This preserves conditions (1) and (2) of the loop invariant.
  \item We then push $i$ onto \texttt{pendingRight} for deferred fixing. This preserves conditions (3) and (4) of the loop invariant.
\end{enumerate}

\paragraph{Termination} The loop processes all values of \texttt{updatedIndices}, including the sentinel index $n$, which is treated as $L$. Substituting $i = n$ in the invariant conditions, we have:
\begin{itemize}
  \item Condition (1) implies that $A[0..\ell-1]$ is sorted and contains its final values.
  \item Condition (2) implies there are no $L$ violations in the range $[\ell, n-1]$.
  \item Conditions (3) and (4) together imply that there are no $R$ violations in the range $[\ell, n-1]$.
  \item Hence we can conclude that the array $A[0..n-1]$ is fully sorted.
\end{itemize}

\figref{fig:delta-sort-example} illustrates this sequence of operations for a sample array.

It is important to note that even though we are doing repeated binary insertions for fixing violations, the overall data movement is reduced as compared to naive independent binary insertions because the binary search range is smaller. To formalize this we need to understand the structure that emerges from Phase~1. If we picture the updated indices arranged in a sequence of $L$ and $R$ labels, based on their direction when first encountered during Phase~2, we can partition them into \emph{segments}.

\begin{definition}[Segment]
\label{def:segment}
A \emph{segment} is a subarray of $A$ spanning indices $[i, j]$, where $0 \le i \le j < n$, such that:
\begin{enumerate}
  \item Either $i = 0$, or $i \in U$ and has direction label $R$.
  \item Either $j = n - 1$, or $j \in U$ and has direction label $L$.
  \item There exist no updated indices $p, q \in U$ with $i < p < q < j$ such that $p$ is $L$ and $q$ is $R$.
  \item There exist no updated indices $p < i$ and $q > j$ such that $(p, q)$ is also a segment.
\end{enumerate}
\end{definition}

\input{\figdir/segment-structure}

More intuitively, a segment is a maximal contiguous region of the array in which all $R$-direction updates precede all $L$-direction updates. The first two conditions define valid segment boundaries, the third enforces the $R^\ast L^\ast$ structure within a segment, and the final condition ensures that segments are maximal and non-overlapping. \figref{fig:segment-structure} illustrates this structure.

\paragraph{Key Insight: Segments can be fixed locally and independently.}

Unlike naive BIS, where each insertion can be at any point in the array, in DeltaSort, \emph{an updated value in a segment cannot cross its segment boundary}. We formally prove this property next.

\begin{lemma}[Movement Confinement]
\label{lem:confinement}
During Phase~2, all value movement is confined within segment boundaries.
\end{lemma}

\begin{proof}
Let $S$ be a segment with $R$-indices $r_0,\ldots,r_{r-1}$ followed by $L$-indices $\ell_0,\ldots,\ell_{l-1}$, where $r+l \ge 1$. After Phase~1, updated values are ordered: $A[r_0] \le \cdots \le A[r_{r-1}] \le A[\ell_0] \le \cdots \le A[\ell_{l-1}]$. An $R$-value cannot have crossed $\ell_0$ (if it exists), and an $L$-value cannot have crossed $r_{r-1}$ (if it exists), since the above ordering must be preserved. Therefore, no value exits its segment.
\end{proof}

\begin{remark}
  Note that \lemref{lem:confinement} also implies \emph{segment independence}. Because no value exits its segment, segment lengths do not change. As a result, movement in a segment does not interfere with movement in another segment, which implies segments can be fixed independently in any order. This opens opportunities for \textbf{parallelization}, where segments could be fixed concurrently. We leave this investigation to future work.
\end{remark}

\subsection{Complexity Analysis}
\label{sec:complexity}

We analyze the expected total data movement incurred during Phase~2 of DeltaSort under a random bounded-range update model. In this model, updated values are drawn uniformly at random from a \emph{fixed} value range. Choosing values at random avoids introducing any special structure that could bias the analysis.

\begin{remark}[Choice of Update Model]
\label{rem:update-model}
Several update models could be considered. For example, a \emph{bounded rank displacement} model constrains updates such that for every updated value $|Rank_{\text{new}} - Rank_{\text{old}}| \le b$ for some bound $b$, while a \emph{clustered update} model restricts updated indices to a region with $u_{\max} - u_{\min} \le c$. Different models interact with DeltaSort’s structure in different ways: bounded displacement directly limits movement and is therefore favorable, whereas clustered updates may either help or hinder performance depending on the induced directional pattern. In this work, we adopt a \emph{random update model as a neutral baseline}. Specifically, updated indices are chosen uniformly at random from $\{0, \ldots, n-1\}$, and updated values are drawn independently from a fixed range of permissible values. This model introduces no additional structure that the algorithm can exploit, nor does it adversarially bias updates toward worst-case configurations (see~\figref{fig:worst-case} for an illustration). Analyzing DeltaSort under more structured update models is left to future work.
\end{remark}

\begin{lemma}[Expected Movement]
\label{lem:movement-bound}
Under the random bounded-range update model, for an array of size $n$ with $k$ updated indices, the expected total data movement incurred during DeltaSort's Phase~2 is $O(n \sqrt{k})$.
\end{lemma}

\begin{proof}
After Phase~1, we have two sorted sequences of $k$ values each:
\begin{itemize}
  \item \textbf{Indices:} The $k$ updated positions $i_0 < i_1 < \cdots < i_{k-1}$, sorted.
  \item \textbf{Values:} The $k$ updated values $v_0 \le v_1 \le \cdots \le v_{k-1}$, sorted.
\end{itemize}
Phase~1 pairs these by rank: the $j$-th smallest value $v_j$ is placed at the $j$-th smallest index $i_j$. Since both indices and values are drawn uniformly at random, each sorted sequence forms an \emph{order statistic} of $k$ uniform samples.

Although values may be drawn from a different range than indices, what matters for determining direction is the value's \emph{target position}---where it belongs in the final sorted array. For a value $v_j$ in a sorted array of $n$ values spanning range $[0, M)$, its target position is approximately $t_j = v_j \cdot n / M$. Since $v_j$ is uniform over $[0, M)$, the target position $t_j$ is uniform over $[0, n)$---the same range as the indices. Thus, we can analyze both sequences as order statistics of uniform samples from the same range.

Define the \emph{gap} at rank $j$ as:
\[
g_j = t_j - i_j
\]
This measures how far ahead (positive) or behind (negative) the value's target position is relative to its current position. By~\defref{def:direction}:
\begin{itemize}
  \item $g_j < 0$ implies the value is too small for its position, so direction is $L$.
  \item $g_j \ge 0$ implies the value is adequate or too large, so direction is $R$.
\end{itemize}

The gap evolves as we move through the ranks. Let $\Delta i_j = i_{j+1} - i_j$ and $\Delta t_j = t_{j+1} - t_j$ denote the spacings between consecutive order statistics. The change in gap is:
\[
g_{j+1} - g_j = \Delta t_j - \Delta i_j
\]
Since both $\Delta t_j$ and $\Delta i_j$ are spacings of independent uniform order statistics, they have the same expected value ($n/(k+1)$) and are independent of each other. Therefore, the increment $g_{j+1} - g_j$ has mean zero and behaves as a random step. This means the sequence $g_1, g_2, \ldots, g_k$ evolves as a \emph{symmetric random walk}.

By~\defref{def:segment}, segment boundaries occur at $L \to R$ transitions---precisely when the gap $g_j$ crosses from negative to non-negative. This corresponds to an \emph{upcrossing} of zero in the random walk. A classical result from random walk theory~\cite{feller1968} states that a symmetric random walk of $k$ steps has expected number of zero crossings $\Theta(\sqrt{k})$. Since $L \to R$ transitions account for roughly half of all zero crossings, the expected number of segments is $O(\sqrt{k})$.

With $O(\sqrt{k})$ segments partitioning the array of size $n$, the expected width of each segment is $O(n / \sqrt{k})$. By~\lemref{lem:confinement}, movement is confined within segments. Since each of the $k$ updated values moves at most the width of its containing segment, the expected total movement is:
\[
\mathbb{E}[\text{total movement}] \le k \cdot O\!\left(\frac{n}{\sqrt{k}}\right) = O(n\sqrt{k}) \qedhere
\]
\end{proof}

\begin{remark}[Worst Case Movement]
While the expected movement is $O(n\sqrt{k})$, the worst-case movement is $O(nk)$. This occurs when updated values form a single segment spanning the entire array. This happens when updates \emph{cluster monotonically} at the start or end of the array (see~\figref{fig:worst-case}). Hence, in a practical setting, a hybrid algorithm will need to fall back to ESM or full re-sort based on the number of segments.
\end{remark}

\input{\figdir/worst-case}

\begin{lemma}[Comparison Count]
\label{lem:comparison-count}
DeltaSort performs $O(k \log n)$ comparisons.
\end{lemma}

\begin{proof}
Phase~1 sorts the $k$ updated values, requiring $O(k \log k)$ comparisons. In Phase~2, each updated value is fixed using binary search within its containing segment. As shown in the proof of~\lemref{lem:movement-bound}, the expected number of segments is $O(\sqrt{k})$, so the expected width of a segment is $O(n/\sqrt{k})$. Since searches are confined within segment boundaries, each fix requires $O(\log (n/\sqrt{k}))$ comparisons, for a total of $O(k \log (n/\sqrt{k}))$ comparisons in Phase~2. Summing both gives:
\[
O(k \log k) + O(k \log (n/\sqrt{k})) = O\!\bigl(k(\log k + \log n - \tfrac{1}{2}\log k)\bigr) = O(k \log (n \sqrt{k})) = O(k \log n) \qedhere
\]
\end{proof}

\begin{theorem}[Time Complexity]
\label{thm:time-complexity}
DeltaSort runs in $O(n\sqrt{k})$ expected time.
\end{theorem}

\begin{proof}
By~\lemref{lem:movement-bound}, the expected total data movement is $O(n\sqrt{k})$. By~\lemref{lem:comparison-count}, the total number of comparisons is $O(k \log n)$. Phase~1 also contributes $O(k \log k)$ movement for sorting $k$ values. Since $k \log n = o(n\sqrt{k})$ for $k = o(n)$, the total is $O(n\sqrt{k})$.
\end{proof}

\begin{theorem}[Space Complexity]
\label{thm:space}
DeltaSort uses $O(k)$ auxiliary space.
\end{theorem}

\begin{proof}
Phase~1 sorts $k$ values using $O(k)$ space. Phase~2 maintains a pending stack of at most $k$ indices. Therefore, the total auxiliary space is $O(k)$.
\end{proof}

\subsection{Empirical Validation}
\label{sec:empirical-validation}

\figref{fig:complexity-analysis} presents empirical validation of DeltaSort's complexity bounds across multiple scales of $n$ and $k$. For movement complexity, we normalize the measured movement by $n\sqrt{k}$; for comparison complexity, we normalize by $k\log(n\sqrt{k})$. If the theoretical bounds are tight, these normalized values should be approximately constant across all scales.

The results confirm both bounds: movement normalized by $n\sqrt{k}$ converges to approximately~$0.4$ across four orders of magnitude of~$n$, and comparisons normalized by $k\log(n\sqrt{k})$ converge to approximately~$1.4$. The overlap of curves for different~$n$ validates that the asymptotic analysis correctly captures the algorithm's behavior.

\input{\figdir/complexity-analysis}