%==============================================================================
\section{Conclusion}
\label{sec:conclusion}
%==============================================================================

In this paper, we studied the problem of maintaining sorted arrays under incremental \emph{known} updates. We argued that this \emph{update-aware} setting arises naturally in many practical systems. Within this model, we presented \emph{DeltaSort}, an incremental sorting algorithm that batches multiple updates instead of applying them independently. The key idea is to exploit structure created by pre-sorting updated values, which induces segments and enables localized, non-overlapping fixes. This enables DeltaSort to reduce redundant data movement. Our theoretical analysis shows that DeltaSort matches the comparison efficiency of BIS while exhibiting $O(n \sqrt k)$ data movement under the random bounded-range update model. Experimental results in Rust demonstrate that DeltaSort outperforms both BIS and ESM for small update batches, which is the most practically relevant operating range for incremental workloads. For larger update batches, ESM's linear merge pass becomes preferable. To stress-test the underlying cost assumptions, we also evaluated DeltaSort in the V8 runtime, where memory movement is less predictable. In this setting, DeltaSort remains faster than BIS by $\sim$2--3$\times$ but the two algorithms exhibit similar growth rates, and as a result have much lower crossover thresholds, highlighting that DeltaSort's full complexity advantage depends on execution environments with transparent and contiguous memory layouts.

More broadly, this work suggests a shift in how sorting primitives can be designed in systems that already track updates explicitly. Traditional adaptive sorting algorithms, such as TimSort~\cite{timsort}, \emph{discover} structure by observing runs and local order in the input. DeltaSort explores a complementary direction: when update metadata is available, structure need not be discovered—it can be \emph{preserved} and used directly. This perspective aligns with modern dataflow and incremental processing systems~\cite{kleppmann2017ddia,wikipediaCDC}, which already propagate updates at higher levels of the stack to avoid redundant computation. By extending update-awareness down to low-level ordering primitives, we open a new space of algorithmic trade-offs that lie beyond what blind, structure-discovery-based sorting algorithms can achieve. This work demonstrates that even for heavily optimized primitives like sorting, explicit update-awareness enables fundamentally different algorithmic choices and warrants further exploration.