%==============================================================================
\section{Future Work}
\label{sec:future}
%==============================================================================

This work opens up several directions for further investigation.

First, while our analysis focuses on a bounded-range update model with uniformly random updates, many practical workloads exhibit additional structure. Examples include bounded-window updates, value perturbation models, and temporally correlated updates. These models may further reduce the expected movement and comparison costs of DeltaSort, potentially expanding the regime in which it outperforms existing techniques. A more systematic study of such update models would help characterize when update-aware sorting provides the greatest benefit.

Second, DeltaSort's repair process naturally decomposes the array into independent segments. This segmentation suggests opportunities for parallel execution, where different segments could be repaired concurrently without interference. In contrast, Binary-Insertion-Sort is inherently sequential due to overlapping in-place shifts, and Extract–Sort–Merge parallelizes only partially while incurring substantial auxiliary space costs. Exploring parallel variants of DeltaSort and understanding their scalability on multi-core and NUMA systems is a promising direction for future work.

Third, our experimental results reveal that crossover thresholds between update-aware algorithms and full re-sorting depend on array size, execution environment, and memory behavior. In particular, we observe a decline in DeltaSort's advantage for very large arrays, likely due to cache effects and data movement overheads. A deeper investigation into these effects, including cache-aware and memory-bandwidth-aware models, could lead to improved implementations or refined hybrid strategies.

Finally, while this paper treats DeltaSort as a standalone algorithm, practical systems will benefit from adaptive hybrid strategies that select among Binary-Insertion-Sort, DeltaSort, Extract–Sort–Merge, and full re-sorting based on observed update sizes and system constraints. Designing heuristic-based, low-overhead mechanisms for such dynamic selection remains an open challenge.
