%==============================================================================
\section{Introduction}
\label{sec:introduction}
%==============================================================================

Sorting is among the most heavily optimized primitives in modern systems, backed by decades of deep research. Standard library implementations---TimSort~\cite{timsort}, Introsort~\cite{10.5555/261387.261395}, and DriftSort~\cite{driftsort} deliver excellent performance for general inputs by exploiting existing order, cache locality, and adaptive strategies. However, these algorithms operate under a \emph{blind} model: they discover structure dynamically rather than being explicitly informed about which values have been updated since the previous sort. While this model is more general and makes things straightforward for the caller, it also \emph{misses opportunities for optimization when such information is available}. Modern systems often maintain metadata about updated records as part of their execution pipeline~\cite{kleppmann2017ddia,wikipediaCDC}, making it possible to expose this information to lower-level sorting primitives. In the absence of a primitive that can accept this as a first-class input, they must either perform a full re-sort or use standard insertion-based or merge-based incremental techniques.

\input{\figdir/incremental-sorting-algorithms}

As an example, consider a modern client application that needs to display a large dynamic sorted list of items (e.g., task management applications or interactive dashboards). Modern applications have tight latency budgets for main-thread execution, beyond which users perceive the interface as unresponsive \cite{webdev_inp,card1983psychology}. As a result, UI frameworks commonly encourage memoization to avoid repeated recomputation of expensive derived values such as sorted views \cite{react_usememo}. However, such memoization is typically coarse-grained: cached results are invalidated whenever dependencies change, without exploiting knowledge of which positions were updated. Consequently, repeated full re-sorting can occupy the main thread long enough to exceed responsiveness budgets, leading to janky user experiences \cite{webdev_inp,googleRAIL}. If a sorting primitive could be informed of which values were updated, it could potentially restore sorted order more efficiently than a full re-sort, enabling more responsive user interfaces.

Once the indices of the updates are known, how do we exploit this extra information to restore order more efficiently than performing a full re-sort? While similar questions arise for a variety of ordered data structures (e.g., B-Trees, heaps and BSTs), in this paper we deliberately focus on \emph{arrays}, which provide a contiguous memory layout. This choice is intentional: arrays remain the dominant representation for sorted data in performance-critical systems because they offer superior cache locality and enable SIMD/vectorized processing~\cite{hennessy2017computer}—benefits that pointer-based structures fundamentally cannot provide. As a result, many systems prefer to preserve array layouts even when updates are frequent, rather than switching to fully dynamic structures~\cite{drepper2007memory,10.1145/1376616.1376712,boncz2005monetdb}. This also allows us to cleanly isolate the algorithmic consequences of exposing update indices to the sorting routine and study the resulting trade-offs in a concrete setting.

Given an array of size $n$, with $k$ updates, existing techniques fall into two broad categories:
\begin{enumerate}
  \item \textbf{Insertion-based}: Each updated value is inserted into its correct position one by one. We use Binary-Insertion-Sort (\emph{BIS}) as a representative of this category which uses binary search to find the correct position for each updated value. This approach uses only $O(1)$ auxiliary space but incurs $O(nk)$ data movement, making it suitable only when $k \ll n$.
  \item \textbf{Merge-based}: Updated values are extracted into a separate buffer, sorted using an efficient $O(k \log k)$ algorithm, and then merged back into the original array. We use Extract–Sort–Merge (\emph{ESM}) as a representative of this category, which compacts the unchanged elements and merges the updates values from backward. It runs in $O(k \log k + n)$ time using only $O(k)$ auxiliary space, but always touches the whole array irrespective of $k$.
\end{enumerate}

So the optimal strategy is: Use BIS when $k \ll n$, ESM otherwise. This raises the question: \emph{are there algorithms that can outperform these approaches} for practically relevant workloads? In this paper, we introduce such an algorithm. Specifically, this paper makes the following contributions:

\begin{enumerate}
  \item \textbf{Update-aware sorting model:} We formulate a sorting model in which the sorting routine is explicitly informed of the updated indices since the previous sort. Under this model, BIS and ESM are the best known baseline algorithms\footnote{Other potential baselines either reduce to trivial combinations of BIS and ESM, or rely on in-place merge algorithms~\cite{kronrod1969optimal} which don't naturally fit our problem model. We evaluated various other baselines to ensure we picked the best performing baselines from both categories (see source repository~\cite{deltasort-repo} for benchmarks).} that naturally exploit knowledge of which indices have been updated.

  \item \textbf{DeltaSort algorithm:} We present \emph{DeltaSort}, an update-aware insertion-based incremental sorting algorithm that reduces data movement significantly by identifying \emph{independent regions}, and restricting movement within these regions. Although we study DeltaSort as a standalone algorithm for clarity, for practical application, it is best viewed as a \emph{building block} in a hybrid strategy that combines multiple techniques.
  
  \item \textbf{Theoretical analysis:} Under a random bounded-range update model, we show that DeltaSort achieves $O(n\sqrt{k})$ expected time complexity using $O(k)$ auxiliary space.
  
  \item \textbf{Experimental validation:} We benchmark Rust implementations of DeltaSort, BIS, and ESM on synthetic datasets. In our evaluation, DeltaSort is the \emph{fastest algorithm for small update batches} ($k \lesssim 1\%$ for $n=100K$), which is the most practically relevant range for incremental workloads.
\end{enumerate}

\input{\figdir/algorithm-comparison}
