\documentclass[11pt]{article}

\usepackage{amsmath,amssymb}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{geometry}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{tikz}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}
\usetikzlibrary{positioning,arrows.meta,patterns,calc,decorations.pathreplacing}

\geometry{margin=1in}

% Theorem environments
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

\title{DeltaSort: Fast incremental repair of sorted arrays}
\author{Shubham Dwivedi \\
\small Independent Researcher \\
\small \texttt{shubd3@gmail.com}
}
\date{December 2025}

\begin{document}
\maketitle

\begin{abstract}
Maintaining sorted order under incremental updates is fundamental in read-heavy systems.
When $k$ out of $n$ elements change, standard approaches either re-sort entirely in
$O(n \log n)$ time or perform $k$ independent binary insertions with $O(kn)$ worst-case
element movement.

This paper presents \emph{DeltaSort}, a coordinated repair algorithm that exploits knowledge of
which indices changed. The key insight is a \textbf{pre-sorting phase} that establishes
monotonicity among dirty values, enabling \textbf{progressive search narrowing} and
\textbf{movement cancellation}---when dirty values would ``cross'' (one moving left,
another right over the same positions), pre-sorting reassigns them to minimize displacement.
Combined with \textbf{direction-aware processing}, DeltaSort achieves $O(k \log n)$
comparisons---optimal for comparison-based algorithms.

Correctness is proven via loop invariants and validated through extensive randomized testing.
Benchmarks using a Rust implementation show DeltaSort achieves 5--17$\times$ speedup over
native sort for delta sizes up to 20--30\% of the array, with the crossover point (where
native sort becomes faster) occurring at approximately 25\% dirty elements across array
sizes from 1K to 1M.
\end{abstract}

%==============================================================================
\section{Introduction}
%==============================================================================

Sorting is among the most heavily optimized primitives in modern systems. Standard library
implementations---TimSort~\cite{timsort}, Introsort~\cite{musser1997introspective}, and
PDQSort~\cite{peters2021pdqsort}---deliver excellent performance for general inputs.
However, these algorithms operate under a \emph{blind} model: they discover structure
rather than being informed which elements changed.

In many production systems, this assumption is unnecessarily pessimistic:
\begin{itemize}
  \item \textbf{UI frameworks} tracking which rows in a sorted table were edited
  \item \textbf{Game leaderboards} where player scores update sparsely
  \item \textbf{Database indices} rebuilt after targeted UPDATE statements
  \item \textbf{Real-time dashboards} with streaming metric updates
\end{itemize}

These systems know exactly which indices changed, yet typically discard this information.

\paragraph{The Efficiency Problem.}
The standard approach---$k$ independent binary insertions---is correct but has efficiency
limitations:

\begin{enumerate}
  \item \textbf{Full-range search}: Each insertion searches the entire array ($O(\log n)$
        comparisons) even when dirty elements cluster in a small region.
  
  \item \textbf{Redundant movement}: Each insertion may shift $O(n)$ elements. When two
        dirty elements would move in opposite directions across the same clean elements,
        those clean elements shift multiple times---first one way, then back.
\end{enumerate}

\paragraph{Contributions.}
This paper makes the following contributions:

\begin{enumerate}
  \item \textbf{DeltaSort Algorithm} (\S\ref{sec:algorithm}): A three-phase algorithm that
        coordinates repairs through pre-sorting, direction classification, and progressive
        search narrowing.
  
  \item \textbf{Movement Cancellation} (\S\ref{sec:algorithm}): Pre-sorting is shown to
        eliminate redundant movement when dirty values would otherwise ``cross.''
  
  \item \textbf{Correctness Proof} (\S\ref{sec:correctness}): Formal proof via loop invariants
        that DeltaSort produces correctly sorted output.
  
  \item \textbf{Empirical Validation} (\S\ref{sec:experiments}): Extensive randomized testing
        across array sizes, delta volumes, and comparator costs.
\end{enumerate}

%==============================================================================
\section{Related Work}
\label{sec:related}
%==============================================================================

\paragraph{Adaptive Sorting.}
Algorithms like TimSort~\cite{timsort} and natural merge sort~\cite{knuth1998art} exploit
existing runs. Mannila~\cite{mannila1985measures} formalized presortedness measures.
However, these algorithms must \emph{discover} structure through $\Omega(n)$ scans;
dirty indices are assumed to be \emph{given}.

\paragraph{Dynamic Data Structures.}
Self-balancing trees (AVL~\cite{avl1962}, red-black~\cite{guibas1978dichromatic},
B-trees~\cite{bayer1970organization}) and skip lists~\cite{pugh1990skip} support
$O(\log n)$ updates but sacrifice contiguous memory layout.

\paragraph{Library Sort.}
Bender et al.~\cite{bender2006insertion} leave gaps for $O(\log n)$ amortized insertions,
but address a different problem (online insertion) with space overhead.

\paragraph{Contribution Positioning.}
This work uniquely combines: (1) array semantics, (2) batched updates with known dirty
indices, and (3) coordinated processing that enables bounded search ranges.

%==============================================================================
\section{Problem Model}
\label{sec:model}
%==============================================================================

\begin{definition}[Informed Incremental Sorting]
\label{def:problem}
Given:
\begin{itemize}
  \item Array $A[0..n-1]$ previously sorted under comparator $\texttt{cmp}$
  \item Dirty indices $D = \{d_1, \ldots, d_k\}$ where values were modified
  \item Clean elements (indices not in $D$) retain their original values
\end{itemize}
Restore $A$ to sorted order with respect to $\texttt{cmp}$.
\end{definition}

\begin{definition}[Clean and Dirty Elements]
Element $A[i]$ is \emph{dirty} if $i \in D$; otherwise it is \emph{clean}.
\end{definition}

\begin{definition}[Monotonicity Violation]
\label{def:monotonicity-violation}
Two dirty indices $d_i, d_j$ with $d_i < d_j$ exhibit a \emph{monotonicity violation}
if their current values satisfy $\texttt{cmp}(A[d_i], A[d_j]) > 0$, i.e., the element at
the lower index is greater than the element at the higher index.
\end{definition}

\begin{definition}[Pass-Over Count]
For a position $c$, define the \emph{pass-over count} $M_c$ as the number of dirty
elements whose movement ``passes over'' position $c$:
\[
M_c = |\{d \in D : \min(d, \pi(d)) \leq c < \max(d, \pi(d))\}|
\]
where $\pi(d)$ is the final position of the dirty element originally at index $d$.
\end{definition}

\subsection{Baseline Algorithms}

\paragraph{Native Sort.}
Re-sort the entire array: $O(n \log n)$ comparisons, $O(n \log n)$ movements.

\paragraph{Binary Insertion (BI).}
For each $d \in D$: extract all dirty values, then for each: binary search for correct
position, reinsert. Cost: $O(k \log n)$ comparisons, $O(kn)$ worst-case movement.
Always correct, but searches the full array range for each insertion.

\paragraph{Extract-Sort-Merge (ESM).}
Extract dirty values, sort them, merge with clean elements.
Cost: $O(k \log k + n)$ comparisons, $O(n)$ movement.
Always correct but requires $O(n)$ auxiliary space.

\subsection{Lower Bounds}

\begin{proposition}[Comparison Lower Bound]
\label{prop:comparison-lb}
Any comparison-based algorithm requires $\Omega(k \log n)$ comparisons in the worst case.
\end{proposition}

\begin{proof}
Each dirty element can occupy any of $n$ final positions; $n^k$ configurations require
$\log_2(n^k) = k \log n$ comparisons to distinguish.
\end{proof}

\begin{proposition}[Movement Lower Bound]
\label{prop:movement-lb}
For any incremental repair algorithm (one that moves elements one at a time rather than
using auxiliary arrays), the clean element at position $c$ must shift at least $M_c$ times.
\end{proposition}

\begin{proof}
Consider the clean element initially at position $c$. Its final position is
$c' = c - L_c + R_c$, where $L_c$ is the number of dirty elements moving from right of
$c$ to left of $c$, and $R_c$ is the number moving from left to right of $c$.

In any incremental repair, when dirty element $d$ moves from position $d$ to $\pi(d)$:
\begin{itemize}
  \item If $d > c$ and $\pi(d) \leq c$: elements in $[\pi(d), d-1]$ shift right, including $c$.
  \item If $d < c$ and $\pi(d) \geq c$: elements in $[d+1, \pi(d)]$ shift left, including $c$.
\end{itemize}
Each of the $M_c = L_c + R_c$ dirty elements passing over $c$ causes exactly one shift.
\end{proof}

\begin{remark}
A merge-based algorithm (like ESM) avoids this by using $O(n)$ auxiliary space to place
each element directly at its final position. The lower bound applies to in-place
incremental approaches.
\end{remark}

%==============================================================================
\section{DeltaSort Algorithm}
\label{sec:algorithm}
%==============================================================================

\subsection{Overview}

DeltaSort operates in three phases:
\begin{enumerate}
  \item \textbf{Phase 1 (Preparation):} Extract dirty values, sort them, write back to
        dirty positions in index order. This establishes monotonicity.
  \item \textbf{Phase 2 (Coordinated Repair):} Process dirty indices left-to-right,
        immediately repairing LEFT violations while deferring RIGHT violations.
  \item \textbf{Phase 3 (Flush):} Process deferred RIGHT violations right-to-left.
\end{enumerate}

\begin{figure}[t]
\centering
\begin{tikzpicture}[
    element/.style={minimum width=0.7cm, minimum height=0.7cm, draw, font=\small},
    dirty/.style={element, fill=red!25},
    clean/.style={element, fill=gray!15},
    arrow/.style={-{Stealth}, thick, blue!70!black},
    phase/.style={font=\small\bfseries, anchor=west}
]

% Phase labels
\node[phase] at (-2.5, 0) {Initial:};
\node[phase] at (-2.5, -1.3) {Phase 1:};
\node[phase] at (-2.5, -2.6) {Phase 2-3:};
\node[phase] at (-2.5, -3.9) {Final:};

% Initial: sorted array [1,3,5,7,9] with dirty at 1,3 getting values 8,2
\node[clean] (i0) at (0*0.8, 0) {1};
\node[dirty] (i1) at (1*0.8, 0) {8};
\node[clean] (i2) at (2*0.8, 0) {5};
\node[dirty] (i3) at (3*0.8, 0) {2};
\node[clean] (i4) at (4*0.8, 0) {9};
\node[font=\tiny, below=0.1cm of i1] {$d_1$};
\node[font=\tiny, below=0.1cm of i3] {$d_2$};

% After Phase 1: sort {8,2} -> {2,8}, write to {1,3}
\node[clean] (p10) at (0*0.8, -1.3) {1};
\node[dirty] (p11) at (1*0.8, -1.3) {2};
\node[clean] (p12) at (2*0.8, -1.3) {5};
\node[dirty] (p13) at (3*0.8, -1.3) {8};
\node[clean] (p14) at (4*0.8, -1.3) {9};
\node[font=\tiny, text=gray] at (2.5*0.8, -0.65) {sort \{8,2\}$\to$\{2,8\}};

% After Phase 2-3: 8 needs to move right
\node[clean] (p20) at (0*0.8, -2.6) {1};
\node[dirty] (p21) at (1*0.8, -2.6) {2};
\node[clean] (p22) at (2*0.8, -2.6) {5};
\node[dirty,fill=yellow!40] (p23) at (3*0.8, -2.6) {8};
\node[clean] (p24) at (4*0.8, -2.6) {9};
\draw[arrow] (p23.east) to[out=0,in=180] ++(0.6,0);

% Final
\node[clean] (f0) at (0*0.8, -3.9) {1};
\node[dirty] (f1) at (1*0.8, -3.9) {2};
\node[clean] (f2) at (2*0.8, -3.9) {5};
\node[dirty] (f3) at (3*0.8, -3.9) {8};
\node[clean] (f4) at (4*0.8, -3.9) {9};

\end{tikzpicture}
\caption{DeltaSort example. Dirty indices $\{1,3\}$ with values $\{8,2\}$. Phase 1 sorts
dirty values to $\{2,8\}$ and writes back, establishing monotonicity for bounded search.}
\label{fig:algorithm}
\end{figure}

\subsection{Why Phase 1 Matters: Enabling Bounded Search}

\begin{example}[Benefit of Pre-sorting]
\label{ex:presort}
Consider array $[1, 8, 5, 2, 9]$ with dirty indices $\{1, 3\}$ (original sorted array
was $[1, 3, 5, 7, 9]$, positions 1 and 3 were updated to 8 and 2).

With standard binary insertion (extract-then-insert), the algorithm would:
\begin{itemize}
  \item Extract dirty values $\{8, 2\}$, leaving placeholders or compacting
  \item Insert each back via binary search over the \emph{entire} remaining array
\end{itemize}
This is correct but each insertion searches the full range.
\end{example}

Phase 1 enables a key optimization: sort $\{8, 2\} \to \{2, 8\}$ and write to positions
$\{1, 3\}$, yielding $[1, 2, 5, 8, 9]$. Now dirty elements are in their correct relative
order, allowing DeltaSort to use \textbf{bounded search ranges}---once a LEFT-moving
element settles at position $t$, subsequent LEFT searches need only consider $[t+1, ...]$.

\begin{lemma}[Phase 1 Establishes Monotonicity]
\label{lem:monotonicity}
After Phase 1, for any dirty indices $d_i < d_j$, it holds that $\texttt{cmp}(A[d_i], A[d_j]) \leq 0$.
\end{lemma}

\begin{proof}
Phase 1 sorts dirty values and assigns the $i$-th smallest value to the $i$-th smallest
index. Thus $A[d_i] \leq A[d_j]$ for $d_i < d_j$.
\end{proof}

\begin{remark}[Bounded Search via \texttt{leftBound}]
\label{rem:leftbound}
The implementation maintains a \texttt{leftBound} variable that advances after each
LEFT move. When a dirty element moves from index $i$ to position $t < i$, the algorithm sets
$\texttt{leftBound} = t + 1$. Subsequent LEFT searches only consider $[\texttt{leftBound}, i-1]$,
not $[0, i-1]$. This progressively narrows the search range as LEFT-moving elements
are processed.
\end{remark}

\begin{remark}[Movement Locality]
\label{rem:locality}
Empirical observation suggests that after Phase 1, dirty elements tend to move within
localized regions. When dirty values are ``inverted'' relative to their indices (e.g.,
a large value at a small index and vice versa), Phase 1 reassigns them such that
the resulting movements are shorter or eliminated entirely. A formal characterization
of the expected movement reduction is left to future work.
\end{remark}

The monotonicity established by Phase 1 is the key to DeltaSort's efficiency:
\begin{enumerate}
  \item \textbf{Progressive search narrowing}: The \texttt{leftBound} variable ensures
        that LEFT searches become progressively narrower as elements are processed.
  \item \textbf{STABLE detection}: After pre-sorting, some dirty elements may already
        be in their correct position relative to neighbors, requiring no movement.
  \item \textbf{Movement cancellation}: When dirty values would ``cross'' (one moving left,
        another right over the same positions), Phase 1 reassigns them to indices such
        that the net displacement is minimized or eliminated (see Example~\ref{ex:cancellation}).
\end{enumerate}

\begin{example}[Movement Cancellation]
\label{ex:cancellation}
Consider array $[1, 3, 5, 7, 9]$ with dirty indices $\{1, 3\}$ receiving values $8$ and $2$
respectively, yielding $[1, 8, 5, 2, 9]$.

\textbf{Binary Insertion} (extract-then-insert):
\begin{itemize}
  \item Extract values at indices $3, 1$ (descending): array becomes $[1, 5, 9]$
  \item Insert $8$: finds position $2$, array becomes $[1, 5, 8, 9]$ (9 shifts right)
  \item Insert $2$: finds position $1$, array becomes $[1, 2, 5, 8, 9]$ (5, 8, 9 shift right)
\end{itemize}
Clean element $5$ shifted during extraction and again during insertion.

\textbf{DeltaSort}:
\begin{itemize}
  \item Phase 1: Sort $\{8, 2\} \to \{2, 8\}$, write to indices $\{1, 3\}$: $[1, 2, 5, 8, 9]$
  \item Phase 2-3: Check directions---both elements are now STABLE!
\end{itemize}
Clean element $5$ never moved. The pre-sorting phase ``cancelled'' the crossing movements
by assigning each value to the index where it already belongs.
\end{example}

\begin{figure}[t]
\centering
\begin{tikzpicture}[
    element/.style={minimum width=0.8cm, minimum height=0.6cm, draw, font=\small},
    dirty/.style={element, fill=red!25},
    clean/.style={element, fill=gray!15},
    moved/.style={element, fill=yellow!40},
    arrow/.style={-{Stealth}, thick},
    label/.style={font=\small\bfseries, anchor=east}
]

% Binary Insertion
\node[label] at (-1.5, 0) {BI:};
\node[font=\scriptsize, anchor=east] at (-1.5, -0.4) {Initial};

% Row 1: Initial [1, 8, 5, 2, 9]
\node[clean] at (0*0.9, 0) {1};
\node[dirty] at (1*0.9, 0) {8};
\node[clean] at (2*0.9, 0) {5};
\node[dirty] at (3*0.9, 0) {2};
\node[clean] at (4*0.9, 0) {9};

% Row 2: After extract [1, 5, 9]
\node[font=\scriptsize, anchor=east] at (-1.5, -1.0) {Extract};
\node[clean] at (0*0.9, -1.0) {1};
\node[moved] at (1*0.9, -1.0) {5};
\node[moved] at (2*0.9, -1.0) {9};
\node[element, draw=none] at (3*0.9, -1.0) {};
\node[element, draw=none] at (4*0.9, -1.0) {};
\draw[arrow, blue!60] (2*0.9, -0.35) -- (1*0.9, -0.65);
\draw[arrow, blue!60] (4*0.9, -0.35) -- (2*0.9, -0.65);

% Row 3: Insert 8 [1, 5, 8, 9]
\node[font=\scriptsize, anchor=east] at (-1.5, -2.0) {Insert 8};
\node[clean] at (0*0.9, -2.0) {1};
\node[clean] at (1*0.9, -2.0) {5};
\node[dirty] at (2*0.9, -2.0) {8};
\node[moved] at (3*0.9, -2.0) {9};
\draw[arrow, blue!60] (2*0.9, -1.35) -- (3*0.9, -1.65);

% Row 4: Insert 2 [1, 2, 5, 8, 9]
\node[font=\scriptsize, anchor=east] at (-1.5, -3.0) {Insert 2};
\node[clean] at (0*0.9, -3.0) {1};
\node[dirty] at (1*0.9, -3.0) {2};
\node[moved] at (2*0.9, -3.0) {5};
\node[moved] at (3*0.9, -3.0) {8};
\node[moved] at (4*0.9, -3.0) {9};
\draw[arrow, blue!60] (1*0.9, -2.35) -- (2*0.9, -2.65);
\draw[arrow, blue!60] (2*0.9, -2.35) -- (3*0.9, -2.65);
\draw[arrow, blue!60] (3*0.9, -2.35) -- (4*0.9, -2.65);

% Shift count for BI
\node[font=\scriptsize, text=red!70!black] at (6.5, -1.5) {Element 5:};
\node[font=\scriptsize, text=red!70!black] at (6.5, -2.0) {shifted 3$\times$};

% Separator
\draw[dashed, gray] (-2, -3.7) -- (8, -3.7);

% DeltaSort
\node[label] at (-1.5, -4.4) {DS:};
\node[font=\scriptsize, anchor=east] at (-1.5, -4.8) {Initial};

% Row 1: Initial [1, 8, 5, 2, 9]
\node[clean] at (0*0.9, -4.4) {1};
\node[dirty] at (1*0.9, -4.4) {8};
\node[clean] at (2*0.9, -4.4) {5};
\node[dirty] at (3*0.9, -4.4) {2};
\node[clean] at (4*0.9, -4.4) {9};

% Row 2: After Phase 1 [1, 2, 5, 8, 9]
\node[font=\scriptsize, anchor=east] at (-1.5, -5.4) {Phase 1};
\node[clean] at (0*0.9, -5.4) {1};
\node[dirty] at (1*0.9, -5.4) {2};
\node[clean] at (2*0.9, -5.4) {5};
\node[dirty] at (3*0.9, -5.4) {8};
\node[clean] at (4*0.9, -5.4) {9};

% Arrows showing value reassignment (not movement)
\draw[arrow, green!60!black, dashed] (1*0.9+0.15, -4.75) to[out=-60,in=60] (1*0.9+0.15, -5.05);
\draw[arrow, green!60!black, dashed] (3*0.9-0.15, -4.75) to[out=-120,in=120] (3*0.9-0.15, -5.05);
\node[font=\tiny, text=green!50!black] at (2*0.9, -4.9) {swap values};

% Row 3: Phase 2-3: Both STABLE
\node[font=\scriptsize, anchor=east] at (-1.5, -6.4) {Phase 2-3};
\node[clean] at (0*0.9, -6.4) {1};
\node[dirty] at (1*0.9, -6.4) {2};
\node[clean] at (2*0.9, -6.4) {5};
\node[dirty] at (3*0.9, -6.4) {8};
\node[clean] at (4*0.9, -6.4) {9};
\node[font=\tiny, text=green!50!black] at (1*0.9, -6.9) {STABLE};
\node[font=\tiny, text=green!50!black] at (3*0.9, -6.9) {STABLE};

% Shift count for DS
\node[font=\scriptsize, text=green!50!black] at (6.5, -5.4) {Element 5:};
\node[font=\scriptsize, text=green!50!black] at (6.5, -5.9) {shifted 0$\times$};

\end{tikzpicture}
\caption{Movement cancellation comparison. Binary Insertion (top) extracts then reinserts,
causing element 5 to shift multiple times. DeltaSort (bottom) reassigns values in Phase 1
so that both dirty elements are immediately STABLE---element 5 never moves.}
\label{fig:cancellation}
\end{figure}

\subsection{Detailed Algorithm}

\begin{algorithm}[t]
\caption{DeltaSort}
\label{alg:deltasort}
\begin{algorithmic}[1]
\Require Array $A[0..n-1]$, dirty indices $D$, comparator $\texttt{cmp}$
\Ensure $A$ is sorted
\Statex
\State \textbf{Phase 1: Establish Monotonicity}
\State $\texttt{dirty} \gets \text{sort}(D)$ \Comment{Sort indices ascending}
\State $\texttt{values} \gets [A[d] : d \in \texttt{dirty}]$
\State $\texttt{values} \gets \text{sort}(\texttt{values}, \texttt{cmp})$
\For{$i \gets 0$ \textbf{to} $|\texttt{dirty}| - 1$}
    \State $A[\texttt{dirty}[i]] \gets \texttt{values}[i]$
\EndFor
\Statex
\State \textbf{Phase 2: Process Left-to-Right}
\State $\texttt{stack} \gets []$; $\texttt{leftBound} \gets 0$
\For{$p \gets 0$ \textbf{to} $|\texttt{dirty}| - 1$}
    \State $i \gets \texttt{dirty}[p]$
    \State $d \gets \Call{Direction}{A, i}$
    \If{$d = \texttt{LEFT}$}
        \State \Call{FlushStack}{$\texttt{stack}, i-1$} \Comment{Flush before processing LEFT}
        \State $t \gets \Call{BinarySearchLeft}{A, A[i], \texttt{leftBound}, i-1}$
        \State \Call{Move}{A, i, t}
        \State $\texttt{leftBound} \gets t + 1$
    \Else
        \State $\texttt{stack.push}(i)$
    \EndIf
\EndFor
\Statex
\State \textbf{Phase 3: Flush Remaining}
\State \Call{FlushStack}{$\texttt{stack}, n-1$}
\end{algorithmic}
\end{algorithm}

\begin{algorithmic}[1]
\Function{Direction}{$A$, $i$}
    \If{$i > 0 \land \texttt{cmp}(A[i-1], A[i]) > 0$}
        \State \Return \texttt{LEFT} \Comment{Violates order with left neighbor}
    \ElsIf{$i < n-1 \land \texttt{cmp}(A[i], A[i+1]) > 0$}
        \State \Return \texttt{RIGHT} \Comment{Violates order with right neighbor}
    \Else
        \State \Return \texttt{STABLE}
    \EndIf
\EndFunction
\end{algorithmic}

\vspace{0.5em}

\begin{algorithmic}[1]
\Function{FlushStack}{$\texttt{stack}$, $\texttt{rightBound}$}
    \While{$\texttt{stack} \neq \emptyset$}
        \State $s \gets \texttt{stack.pop}()$ \Comment{Process in LIFO order}
        \If{$\Call{Direction}{A, s} = \texttt{RIGHT}$}
            \State $t \gets \Call{BinarySearchRight}{A, A[s], s+1, \texttt{rightBound}}$
            \State \Call{Move}{A, s, t}
        \EndIf
    \EndWhile
\EndFunction
\end{algorithmic}

\vspace{0.5em}

\begin{algorithmic}[1]
\Function{Move}{$A$, $from$, $to$}
    \State $v \gets A[from]$
    \If{$from < to$}
        \State Shift $A[from+1..to]$ left by one
    \ElsIf{$from > to$}
        \State Shift $A[to..from-1]$ right by one
    \EndIf
    \State $A[to] \gets v$
\EndFunction
\end{algorithmic}

\subsection{Design Rationale}

\paragraph{Why flush before LEFT?}
When a LEFT-moving dirty element is encountered at index $i$, any RIGHT-moving elements
on the stack (with indices $< i$) must be processed first. Otherwise, moving the LEFT
element could shift the RIGHT elements' positions, invalidating their stack indices.

\paragraph{Why LIFO order?}
Stack elements are pushed in ascending index order. LIFO processing (highest index first)
ensures that when the element at index $s$ is moved, elements at lower indices haven't shifted
yet, so their stack indices remain valid.

\paragraph{Why re-check Direction?}
After Phase 1 writes or after flushing, an element's direction may change (a STABLE
element might become RIGHT, or a RIGHT element might become STABLE). The re-check
avoids unnecessary moves.

%==============================================================================
\section{Correctness Proof}
\label{sec:correctness}
%==============================================================================

\begin{definition}[Violation]
A \emph{violation} at index $i$ exists if $\texttt{cmp}(A[i-1], A[i]) > 0$ (for $i > 0$)
or $\texttt{cmp}(A[i], A[i+1]) > 0$ (for $i < n-1$).
\end{definition}

\begin{lemma}[Violations Only at Dirty Boundaries]
\label{lem:violations}
After Phase 1, violations can only exist at boundaries involving dirty indices.
Specifically, a violation at position $i$ implies $i \in D$ or $i+1 \in D$.
\end{lemma}

\begin{proof}
Clean elements retain their values from the original sorted array. For two adjacent
clean elements at positions $i$ and $i+1$, neither has changed, so
$\texttt{cmp}(A[i], A[i+1]) \leq 0$ (sorted order preserved). Violations only occur
where a dirty element is adjacent to another element with which it violates order.
\end{proof}

\begin{lemma}[Move Correctness]
\label{lem:move-correct}
After $\Call{Move}{A, i, t}$ where $t = \Call{BinarySearchLeft/Right}{\ldots}$:
\begin{enumerate}
  \item The element originally at $A[i]$ is now at position $t$
  \item $\texttt{cmp}(A[t-1], A[t]) \leq 0$ (if $t > 0$)
  \item $\texttt{cmp}(A[t], A[t+1]) \leq 0$ (if $t < n-1$ and within search bounds)
\end{enumerate}
\end{lemma}

\begin{proof}
Binary search finds the correct insertion point by comparing with neighbors. The
element is placed where it satisfies order constraints with its new neighbors.
\end{proof}

\begin{lemma}[Stack Index Validity]
\label{lem:stack-validity}
When an index $s$ is popped from the stack and processed, $s$ still refers to the
same element that was pushed.
\end{lemma}

\begin{proof}
Consider when $s$ was pushed: it was a dirty index with direction RIGHT or STABLE.
Between push and pop, two types of operations can occur:

\emph{(a) LEFT repairs}: These move elements at index $i > s$ to positions $t \leq i - 1$.
Since $s < i$, the region $[t, i-1]$ is to the right of $s$; elements at $s$ don't shift.

\emph{(b) RIGHT repairs from stack flushes}: These process indices in LIFO order.
When $s$ is popped, all indices greater than $s$ have already been processed. Their
movements shift elements to the right, in regions $[s'+1, t']$ where $s' > s$. Since
$s < s'$, position $s$ is unaffected.

In both cases, the element at index $s$ hasn't moved, so $s$ remains valid.
\end{proof}

\begin{figure}[t]
\centering
\begin{tikzpicture}[
    cell/.style={minimum width=0.5cm, minimum height=0.5cm, draw, font=\scriptsize},
    clean/.style={cell},
    left/.style={cell, fill=blue!25},
    right/.style={cell, fill=red!25},
    stable/.style={cell},
    dots/.style={cell, fill=gray!8},
    segment/.style={dashed, gray!70, thick},
    label/.style={font=\tiny}
]

% Array cells showing segments with ... to indicate arbitrary spacing
% Pattern: ... L ... L | ... R ... R | ... S | ... L ... L | ... R R ... | ... S ...
\node[dots] (e0) at (0*0.6, 0) {\dots};
\node[left] (d1) at (1*0.6, 0) {L};
\node[dots] (e2) at (2*0.6, 0) {\dots};
\node[left] (d3) at (3*0.6, 0) {L};
% segment boundary
\node[dots] (e4) at (4*0.6, 0) {\dots};
\node[right] (d5) at (5*0.6, 0) {R};
\node[dots] (e6) at (6*0.6, 0) {\dots};
\node[right] (d7) at (7*0.6, 0) {R};
% segment boundary
\node[dots] (e8) at (8*0.6, 0) {\dots};
\node[stable] (d9) at (9*0.6, 0) {S};
% segment boundary
\node[dots] (e10) at (10*0.6, 0) {\dots};
\node[left] (d11) at (11*0.6, 0) {L};
\node[dots] (e12) at (12*0.6, 0) {\dots};
\node[left] (d13) at (13*0.6, 0) {L};
% segment boundary
\node[dots] (e14) at (14*0.6, 0) {\dots};
\node[right] (d15) at (15*0.6, 0) {R};
\node[dots] (e16) at (16*0.6, 0) {\dots};
\node[right] (d17) at (17*0.6, 0) {R};
\node[dots] (e18) at (18*0.6, 0) {\dots};

% Segment dividers - never right beside R
\draw[segment] (3.5*0.6, -0.45) -- (3.5*0.6, 0.7);
\draw[segment] (9.5*0.6, -0.45) -- (9.5*0.6, 0.7);
\draw[segment] (13.5*0.6, -0.45) -- (13.5*0.6, 0.7);

% Segment labels
\node[font=\tiny, anchor=north] at (1.75*0.6, -0.55) {Seg 1};
\node[font=\tiny, anchor=north] at (6.5*0.6, -0.55) {Seg 2};
\node[font=\tiny, anchor=north] at (11.5*0.6, -0.55) {Seg 3};
\node[font=\tiny, anchor=north] at (16*0.6, -0.55) {Seg 4};

% Legend
\node[left, minimum width=0.4cm, minimum height=0.4cm, label=right:{\tiny LEFT}] at (13*0.6, 1.1) {L};
\node[right, minimum width=0.4cm, minimum height=0.4cm, label=right:{\tiny RIGHT}] at (14.3*0.6, 1.1) {R};
\node[stable, minimum width=0.4cm, minimum height=0.4cm, label=right:{\tiny STABLE}] at (15.6*0.6, 1.1) {S};
\node[dots, minimum width=0.4cm, minimum height=0.4cm, label=right:{\tiny Clean}] at (17.2*0.6, 1.1) {\dots};

% Movement arrows showing bounded regions within segments
\draw[-{Stealth}, blue!70, thick] (2*0.6, 0.4) to[out=150,in=30] (0.5*0.6, 0.4);
\draw[-{Stealth}, red!70, thick] (6*0.6, 0.4) to[out=30,in=150] (7.3*0.6, 0.4);
\draw[-{Stealth}, blue!70, thick] (12*0.6, 0.4) to[out=150,in=30] (10.5*0.6, 0.4);
\draw[-{Stealth}, red!70, thick] (16*0.6, 0.4) to[out=30,in=150] (18*0.6, 0.4);

\end{tikzpicture}
\caption{Segmented view of an array with 9 dirty elements after Phase 1. Each dirty element
is classified as LEFT (L), RIGHT (R), or STABLE (S) based on its violation direction.
Consecutive dirty elements of the same direction type form a \emph{segment}, delimited by
vertical dashed lines. Movement within an L segment is confined to that segment alone (arrows
show leftward movement). Movement within an R segment can span into the following L segment
(arrows show rightward movement). Crucially, all movements remain \emph{bounded}---they never
extend beyond the adjacent segment boundary. This boundedness, enabled by Phase 1's monotonicity,
is key to DeltaSort's efficiency. Boxed ``\dots'' represents arbitrarily many clean elements.}
\label{fig:segments}
\end{figure}

\begin{theorem}[Correctness]
\label{thm:correctness}
DeltaSort produces a correctly sorted array.
\end{theorem}

\begin{proof}
By Lemma~\ref{lem:violations}, after Phase 1, violations only involve dirty indices.
Figure~\ref{fig:segments} illustrates how the array can be viewed as segments where
dirty elements move within bounded regions without crossing segment boundaries.
Each dirty index is processed exactly once in Phases 2-3:
\begin{itemize}
  \item If Direction is LEFT: processed immediately with Move
  \item If Direction is RIGHT or STABLE: pushed to stack, later processed (if still RIGHT)
\end{itemize}

By Lemma~\ref{lem:stack-validity}, stack indices remain valid when processed.
By Lemma~\ref{lem:move-correct}, each Move places the element correctly.
After all dirty indices are processed, no violations remain (all were fixed), so
the array is sorted.
\end{proof}

%==============================================================================
\section{Movement Analysis}
\label{sec:movement}
%==============================================================================

\begin{theorem}[Optimal Per-Element Movement]
\label{thm:movement}
The clean element initially at position $c$ shifts exactly $M_c$ times during DeltaSort
execution, where $M_c$ is its pass-over count. No incremental repair algorithm can
achieve fewer shifts.
\end{theorem}

\begin{proof}
\textbf{Upper bound (DeltaSort achieves $M_c$):}

A clean element at position $c$ shifts when a Move operation has $c$ in its movement region.
This occurs exactly when a dirty element's source and target bracket position $c$:
\begin{itemize}
  \item LEFT move from $i > c$ to $t \leq c$: shifts region $[t, i-1]$, including $c$
  \item RIGHT move from $s < c$ to $t \geq c$: shifts region $[s+1, t]$, including $c$
\end{itemize}

Each dirty element contributes at most one shift to position $c$ (it either passes over
$c$ or doesn't). The total is exactly $M_c = L_c + R_c$.

\textbf{Lower bound (any algorithm needs $\geq M_c$):}

By Proposition~\ref{prop:movement-lb}, any incremental repair algorithm must shift
position $c$ at least $M_c$ times.
\end{proof}

\begin{corollary}[Total Movement]
\label{cor:total-movement}
Total movement is $\sum_{c \notin D} M_c$, which is $O(kn)$ worst case but typically
$O(k \cdot \bar{m})$ where $\bar{m}$ is the average dirty element movement distance.
\end{corollary}

\begin{remark}[No Single-Shift Guarantee]
Clean elements can shift multiple times during incremental repair.
Example: dirty indices $\{5, 7\}$ both receive very small values (e.g., both move to
positions 0 and 1). Elements at positions 1, 2, 3, 4 each shift twice---once for each
dirty element passing over them.

This is \textbf{unavoidable}: those elements must move 2 positions right to accommodate
both insertions. No incremental algorithm can do better.
\end{remark}

\begin{remark}[Comparison to ESM]
Extract-Sort-Merge achieves $O(n)$ total movement by using $O(n)$ auxiliary space to
place each element directly at its final position, avoiding the lower bound. DeltaSort
trades movement efficiency for space efficiency ($O(k)$ auxiliary space).
\end{remark}

%==============================================================================
\section{Complexity Analysis}
\label{sec:complexity}
%==============================================================================

\begin{theorem}[Time Complexity]
\label{thm:time}
DeltaSort runs in $O(k \log k + k \log n + M)$ time, where $M$ is total movement.
\end{theorem}

\begin{proof}
\textbf{Phase 1}: Sort $k$ indices: $O(k \log k)$. Sort $k$ values: $O(k \log k)$.
Write back: $O(k)$.

\textbf{Phases 2-3}: Each dirty index: $O(1)$ direction check, $O(\log n)$ binary search.
Total: $O(k \log n)$. Movement: $O(M)$.
\end{proof}

\begin{theorem}[Space Complexity]
\label{thm:space}
DeltaSort uses $O(k)$ auxiliary space.
\end{theorem}

\begin{theorem}[Comparison Optimality]
\label{thm:optimal-comparisons}
DeltaSort's $O(k \log n)$ comparisons match the information-theoretic lower bound.
\end{theorem}

\begin{table}[h]
\centering
\caption{Algorithm complexity comparison.}
\label{tab:complexity}
\begin{tabular}{l c c c c}
\toprule
Algorithm & Comparisons & Movement & Space & Bounded Search? \\
\midrule
Native Sort & $O(n \log n)$ & $O(n \log n)$ & $O(n)$ & N/A \\
Binary Insertion & $O(k \log n)$ & $O(kn)$ & $O(1)$ & No \\
Extract-Sort-Merge & $O(k \log k + n)$ & $O(n)$ & $O(n)$ & N/A \\
\textbf{DeltaSort} & $O(k \log n)$ & $O(kn)$* & $O(k)$ & Yes \\
\bottomrule
\end{tabular}

\vspace{0.3em}
{\small *Worst case; typically $O(k \cdot \bar{m})$ for average movement distance $\bar{m}$.}
\end{table}

%==============================================================================
\section{Experimental Evaluation}
\label{sec:experiments}
%==============================================================================

\subsection{Setup}

\paragraph{Primary Implementation.} Rust 1.75 (release build with optimizations).

\paragraph{Hardware.} Apple M2 Pro, 16GB RAM, macOS 14.

\paragraph{Algorithms.}
\begin{itemize}
  \item \textbf{Native}: Rust's \texttt{sort\_by} (pattern-defeating quicksort)
  \item \textbf{BI}: Binary Insertion (extract-then-insert)
  \item \textbf{ESM}: Extract-Sort-Merge
  \item \textbf{DS}: DeltaSort
\end{itemize}

\paragraph{Data.} User objects with composite key (country, age, name).
Array sizes $n \in \{1\text{K}, 10\text{K}, 50\text{K}, 100\text{K}, 500\text{K}, 1\text{M}\}$.
Dirty counts $k \in \{1, 5, 10, 20, 50, 100, 200, 500, 1000, 2000, 5000, 10000, 20000\}$.
100 iterations per configuration with 95\% confidence intervals reported.

\subsection{Results}

Table~\ref{tab:rust-results} shows timing results for $n = 50,000$ elements.
DeltaSort consistently outperforms all alternatives up to approximately $k = 10,000$
(20\% of array size), achieving speedups of 7--17$\times$ over Native sort in the
sweet spot ($20 \leq k \leq 200$).

Figure~\ref{fig:crossover} visualizes the crossover between DeltaSort and Native sort
on a linear scale. The crossover occurs at $k \approx 13,700$ (27\% of array size),
clearly showing the region where DeltaSort provides substantial speedups.

Figure~\ref{fig:all-algorithms} compares all four algorithms on a log-log scale,
revealing the orders-of-magnitude performance gap between DeltaSort and the baseline
algorithms (BI and ESM) across the full range of $k$ values.

\begin{table}[t]
\centering
\caption{Execution time (\textmu s) for $n = 50,000$ elements. DS vs Best shows speedup
against the fastest alternative (Native, BI, or ESM).}
\label{tab:rust-results}
\begin{tabular}{r r r r r l}
\toprule
$k$ & Native & BI & ESM & DeltaSort & DS vs Best \\
\midrule
1     & 439 & 40 & 188 & \textbf{1} & 62$\times$ faster \\
5     & 749 & 217 & 293 & \textbf{27} & 8$\times$ faster \\
10    & 747 & 525 & 494 & \textbf{44} & 11$\times$ faster \\
20    & 962 & 782 & 655 & \textbf{49} & 13$\times$ faster \\
50    & 1,350 & 1,994 & 1,307 & \textbf{98} & 13$\times$ faster \\
100   & 1,867 & 4,818 & 2,673 & \textbf{109} & 17$\times$ faster \\
200   & 2,560 & 9,022 & 4,785 & \textbf{178} & 14$\times$ faster \\
500   & 3,703 & 22,112 & 11,655 & \textbf{782} & 5$\times$ faster \\
1,000  & 4,253 & 44,481 & 22,480 & \textbf{465} & 9$\times$ faster \\
2,000  & 4,158 & 87,827 & 43,627 & \textbf{837} & 5$\times$ faster \\
5,000  & 4,223 & 211,016 & 102,198 & \textbf{1,751} & 2.4$\times$ faster \\
10,000 & 4,539 & 377,981 & 176,373 & \textbf{3,326} & 1.4$\times$ faster \\
20,000 & \textbf{5,042} & 615,645 & 259,835 & 6,914 & 1.4$\times$ slower \\
\bottomrule
\end{tabular}
\end{table}

% Figure: DeltaSort vs Native (zoomed to show crossover)
\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
    width=0.9\textwidth,
    height=6cm,
    xlabel={Number of dirty elements ($k$)},
    ylabel={Execution time (\textmu s)},
    xmin=0, xmax=21000,
    ymin=0, ymax=8000,
    xtick={0, 5000, 10000, 15000, 20000},
    xticklabels={0, 5K, 10K, 15K, 20K},
    ytick={0, 2000, 4000, 6000, 8000},
    legend pos=north west,
    legend style={font=\small, fill=white, fill opacity=0.8},
    grid=both,
    grid style={line width=0.1pt, draw=gray!30},
    major grid style={line width=0.2pt, draw=gray!50},
]

% Native sort (relatively flat ~4000-5000 µs)
\addplot[color=blue, mark=square*, thick, mark size=2pt] coordinates {
    (1, 439) (5, 749) (10, 747) (20, 962) (50, 1350) (100, 1867) 
    (200, 2560) (500, 3703) (1000, 4253) (2000, 4158) 
    (5000, 4223) (10000, 4539) (20000, 5042)
};

% DeltaSort (grows linearly with k)
\addplot[color=orange, mark=*, thick, mark size=2pt] coordinates {
    (1, 1) (5, 27) (10, 44) (20, 49) (50, 98) (100, 109) 
    (200, 178) (500, 782) (1000, 465) (2000, 837) 
    (5000, 1751) (10000, 3326) (20000, 6914)
};

% Crossover region annotation
\draw[dashed, thick, gray] (axis cs:13672,0) -- (axis cs:13672,7500);
\node[anchor=south, font=\small] at (axis cs:13672,7500) {crossover $\approx$ 13.7K};

% Shade the "DeltaSort wins" region
\fill[green!10, opacity=0.5] (axis cs:0,0) rectangle (axis cs:13672,8000);

\legend{Native Sort, DeltaSort}
\end{axis}
\end{tikzpicture}
\caption{DeltaSort vs.\ Native Sort for $n = 50,000$. The crossover point occurs at 
$k \approx 13,700$ (27\% of $n$). The shaded region indicates where DeltaSort is faster.}
\label{fig:crossover}
\end{figure}

% Figure: All algorithms comparison
\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
    width=0.9\textwidth,
    height=7cm,
    xlabel={Number of dirty elements ($k$)},
    ylabel={Execution time (\textmu s)},
    xmode=log,
    ymode=log,
    log basis x=10,
    log basis y=10,
    xmin=1, xmax=25000,
    ymin=1, ymax=700000,
    legend pos=north west,
    legend style={font=\small, fill=white, fill opacity=0.9},
    grid=both,
    grid style={line width=0.1pt, draw=gray!30},
    major grid style={line width=0.2pt, draw=gray!50},
]

% Native sort
\addplot[color=blue, mark=square*, thick, mark size=2.5pt] coordinates {
    (1, 439) (5, 749) (10, 747) (20, 962) (50, 1350) (100, 1867) 
    (200, 2560) (500, 3703) (1000, 4253) (2000, 4158) 
    (5000, 4223) (10000, 4539) (20000, 5042)
};

% Binary Insertion
\addplot[color=red, mark=triangle*, thick, mark size=2.5pt] coordinates {
    (1, 40) (5, 217) (10, 525) (20, 782) (50, 1994) (100, 4818) 
    (200, 9022) (500, 22112) (1000, 44481) (2000, 87827) 
    (5000, 211016) (10000, 377981) (20000, 615645)
};

% Extract-Sort-Merge
\addplot[color=green!60!black, mark=diamond*, thick, mark size=2.5pt] coordinates {
    (1, 188) (5, 293) (10, 494) (20, 655) (50, 1307) (100, 2673) 
    (200, 4785) (500, 11655) (1000, 22480) (2000, 43627) 
    (5000, 102198) (10000, 176373) (20000, 259835)
};

% DeltaSort
\addplot[color=orange, mark=*, thick, mark size=2.5pt, line width=1.5pt] coordinates {
    (1, 1) (5, 27) (10, 44) (20, 49) (50, 98) (100, 109) 
    (200, 178) (500, 782) (1000, 465) (2000, 837) 
    (5000, 1751) (10000, 3326) (20000, 6914)
};

\legend{Native, BI, ESM, DeltaSort}
\end{axis}
\end{tikzpicture}
\caption{Execution time comparison of all algorithms for $n = 50,000$ (log-log scale).
DeltaSort (orange) is consistently fastest across all tested $k$ values except $k = 20,000$
where Native sort wins. Note the orders-of-magnitude gap between DeltaSort and BI/ESM.}
\label{fig:all-algorithms}
\end{figure}

\subsection{Crossover Analysis}

A key practical question is: at what delta size should one switch from DeltaSort to
Native sort? A binary search was conducted for the crossover point $k_c$ across array
sizes from 1K to 1M elements. Table~\ref{tab:crossover} summarizes the results.

\begin{table}[t]
\centering
\caption{Crossover point $k_c$ where Native sort becomes faster than DeltaSort.}
\label{tab:crossover}
\begin{tabular}{r r r}
\toprule
$n$ & $k_c$ & $k_c / n$ \\
\midrule
1,000 & 235 & 23.5\% \\
2,000 & 469 & 23.4\% \\
5,000 & 1,211 & 24.2\% \\
10,000 & 2,813 & 28.1\% \\
20,000 & 5,782 & 28.9\% \\
50,000 & 13,672 & 27.3\% \\
100,000 & 31,251 & 31.3\% \\
200,000 & 54,688 & 27.3\% \\
500,000 & 105,469 & 21.1\% \\
1,000,000 & 156,251 & 15.6\% \\
\bottomrule
\end{tabular}
\end{table}

The crossover ratio $k_c / n$ ranges from approximately 15\% to 31\%, with most values
falling in the 20--30\% range. This suggests a practical rule of thumb:

\begin{quote}
\emph{Use DeltaSort when fewer than $\sim$25\% of elements are dirty; otherwise use Native sort.}
\end{quote}

\subsection{JavaScript Implementation}

DeltaSort was also implemented in TypeScript running on Node.js v20 (V8 engine). While
the implementation passes all correctness tests, the performance results show higher
variance due to JIT compilation behavior, garbage collection pauses, and other
engine-level effects. The JavaScript benchmarks will be refined in a future revision
to provide more stable measurements. The Rust implementation provides the authoritative
performance characterization.

\subsection{Analysis}

\paragraph{DeltaSort vs.\ Binary Insertion.}
DeltaSort dramatically outperforms BI across all tested configurations, with speedups
ranging from 8$\times$ to over 100$\times$. This is because BI's $O(kn)$ extraction and
insertion costs dominate, while DeltaSort's in-place rotations avoid element extraction
entirely.

\paragraph{DeltaSort vs.\ ESM.}
DeltaSort also outperforms ESM for all tested $k$ values up to 10,000. ESM's $O(n)$
merge pass becomes competitive only when $k$ is very large, and even then DeltaSort
remains faster in these measurements.

\paragraph{DeltaSort vs.\ Native Sort.}
The crossover with Native sort occurs at approximately 20--30\% dirty elements. Below
this threshold, DeltaSort's $O(k \log n)$ complexity and efficient in-place operations
provide substantial speedups. Above this threshold, Native sort's highly optimized
$O(n \log n)$ implementation wins.

%==============================================================================
\section{Discussion}
\label{sec:discussion}
%==============================================================================

\subsection{What DeltaSort Provides}

\begin{enumerate}
  \item \textbf{Progressive search narrowing}: The \texttt{leftBound} optimization
        reduces search ranges as LEFT-moving elements are processed.
  \item \textbf{Movement cancellation}: When dirty values would cross, pre-sorting
        reassigns them to minimize net displacement (Example~\ref{ex:cancellation}).
  \item \textbf{Optimal comparisons}: $O(k \log n)$, matching the lower bound.
  \item \textbf{Optimal per-element movement}: Each clean element shifts $M_c$ times---the
        minimum for any incremental repair.
  \item \textbf{Low space}: $O(k)$ auxiliary space.
  \item \textbf{Substantial speedup}: 5--17$\times$ over native sort for moderate $k$ (up to $\sim$25\% of $n$).
\end{enumerate}

\subsection{What DeltaSort Does NOT Provide}

\begin{enumerate}
  \item \textbf{Single-shift guarantee}: Clean elements may shift multiple times (unavoidable
        for incremental repair)
  \item \textbf{$O(n)$ total movement}: Worst case is $O(kn)$; use ESM if movement dominates
  \item \textbf{Advantage for tiny $k$}: Phase 1 overhead makes DS slightly slower than
        direct BI for $k \leq 5$
  \item \textbf{Advantage for large $k$}: When $k > 0.25n$, native sort is faster
\end{enumerate}

\subsection{Algorithm Selection Guide}

Based on the Rust benchmarks across array sizes from 1K to 1M elements:

\begin{center}
\begin{tabular}{ll}
\toprule
Condition & Recommendation \\
\midrule
$k \leq 5$ & Binary Insertion (skip Phase 1 overhead) \\
$5 < k < 0.25n$ & \textbf{DeltaSort} \\
$k \geq 0.25n$ & Native Sort \\
\bottomrule
\end{tabular}
\end{center}

The crossover point of approximately 25\% is remarkably stable across array sizes,
ranging from 15\% to 31\% in the experiments (Table~\ref{tab:crossover}).

\subsection{Empirical Validation}

Correctness of the implementation has been validated through extensive randomized testing.
The test suite covers:
\begin{itemize}
  \item Array sizes from $10^1$ to $10^4$ elements
  \item Dirty element ratios from 1\% to 80\% of array size
  \item 10 random iterations per configuration
  \item Random dirty index selection and random value assignment
\end{itemize}

Each test verifies that DeltaSort produces output identical to native sort on the same
input. The test suite has been run successfully across thousands of randomized inputs
without failure, providing high confidence in correctness.

%==============================================================================
\section{Future Work}
\label{sec:future}
%==============================================================================

\paragraph{Formal Locality Bounds.}
A conjecture is that Phase 1's monotonicity property constrains the movement of dirty
elements in a way that reduces total element displacement compared to extract-insert
binary insertion. A formal proof characterizing the expected movement reduction---potentially
in terms of the ``inversion distance'' between original dirty values and their indices---would
strengthen the theoretical foundation.

\paragraph{Cache-Aware Analysis.}
Formalize why bounded search ranges help despite unchanged asymptotic complexity.

\paragraph{Block Storage.}
Analyze DeltaSort for B-tree maintenance, where batched updates may reduce node splits.

\paragraph{JavaScript Implementation Refinement.}
The TypeScript/Node.js implementation shows higher variance due to JIT compilation,
garbage collection, and engine-level effects. Future work will characterize and minimize
these effects to provide stable JavaScript benchmarks.

\paragraph{Adaptive Hybrid.}
Runtime selection between DeltaSort and Native sort based on the 25\% crossover threshold,
potentially with dynamic adjustment based on observed performance.

%==============================================================================
\section{Conclusion}
\label{sec:conclusion}
%==============================================================================

This paper presented DeltaSort, a coordinated incremental repair algorithm for sorted arrays.
The key contributions are:

\begin{enumerate}
  \item \textbf{Pre-sorting phase}: Establishing monotonicity among dirty values enables
        progressive search narrowing and movement cancellation.
  \item \textbf{Provably optimal comparisons}: $O(k \log n)$, matching the lower bound.
  \item \textbf{Provably optimal per-element movement}: Each clean element shifts exactly
        $M_c$ times---the information-theoretic minimum for incremental repair.
  \item \textbf{Substantial practical speedup}: 5--17$\times$ over native sort for delta
        sizes up to $\sim$25\% of array size, validated through Rust benchmarks across
        array sizes from 1K to 1M elements.
\end{enumerate}

The crossover point where native sort becomes faster occurs at approximately 25\% dirty
elements, providing a clear decision boundary for practitioners.

Correctness has been validated through extensive randomized testing across array sizes,
delta volumes, and random value distributions.

The broader lesson is that exploiting application-level knowledge (which indices changed)
enables coordination that blind algorithms cannot achieve. DeltaSort demonstrates that
even with identical asymptotic bounds, careful coordination yields substantial practical gains.

%==============================================================================
\bibliographystyle{plain}
\bibliography{refs}
\nocite{*}
\end{document}
